A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-19 doc/paper.tex
More paper work -- user study.
Commit before big paper changes.

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
   \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{pic/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
   \DeclareGraphicsExtensions{.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi

@@

\section{User Study}
Midway through production, we conducted a brief, four-participant user study in order to gauge how well we were capturing the initial
design goals. The study was, in general, a flexible and exploratory
affair: rather than giving the users any specific tasks or pointed
questions, we asked instead that they choose for themselves which trees to explore and report anything that they deemed interesting. 
Specifically, given trees rooted in the Enron dataset and at least 20 to thirty minutes, the participants situated themselves within the persona 
of a consultant asked to evaluate a company's spreadsheet practice, which could be arranged around questions like which functions on which the
employees most depended, which sheets presented the most anomalous or dangerous designs, and so on. Interpretation of "spreadsheet practice" can certainly vary from participant to participant, but since we wanted a range of views and backgrounds despite the persona, we readily accepted
this interpretive wiggle room. \par
We clarified, furthermore, that these notes could be directed either at the quality of the data conveyed by the tool or at the tool's conveyance
itself. Since the tool's philosophy was based in exploration, we
decided that it would not have been appropriate to restrict what people
were to look for, letting them find interesting stories of formula 
construction for themselves. Additionally, either kind of comment -- on data or on tool -- would direct us back to the central evaluation; a user's
silence, or lack of any interesting findings, would tacitly indicate that
perhaps the tool does not convey information well enough. \par 
Still, several limitations cropped up during the study that complicate
the evaluation. For one thing, though the source of the data was not 
explicitly stated at the outset of the session, several users studied the file names (unchanged as they were) to deduce its roots in finance, if not
Enron itself. Because of these notorious connotations, then, some were
inclined from this moment of discovery to orient their exploration around finance-related functions, potentially limiting their findings or perceived
users for the tool -- though some also said that, without this essential 
context, they could not have properly explored a dataset in the first place. Comments like these point out the trade-offs of unguided exploration, too: though it might not restrict them to a single purpose, it
might, at the same time, not afford them any mindset at all with which to 
interpret the data. Another potential problem is that the data which users
explored was processed and created before we had fully completed the tool,
meaning some of the counts and patterns in the trees were inaccurate. 
However, we decided that this was not a pressing concern, being that we 
were primarily with whether any interesting reports could be made at all, 
not with whether these reports were 100\% accurate assessments of reality. \par
Nevertheless, these user studies were a formative moment in evaluating
the tool's performance, generating a number of directions or corrections
in the process.
\section{Case Study}
The obvious question to pose to any visualization is this: Why does it matter that we see this way? In other words, it is not enough to prove that something can be visualized; instead we must also demonstrate what we can be gained from any particular visualization. As such, we raised a few tasks in the introduction in which we thought the tool could help: identifying places for API improvement, detecting bad smells or common programmer misunderstanding, and guiding spreadsheet education. Thus, to demonstrate the tool's applicability, we sought out places in the tree which best exemplify these respective concerns.

\subsection{Bad Smells}
Fowler's (?) seminal description of code smells underscored
an important point of code quality: between perfect code and 
bug-crippled spaghetti,  there is a spectrum of code designs 
which, by themselves, are not faulty but nevertheless may
point to problems and vulnerabilities in how code is written.
Code smells, as we call them, soon commanded attention and earned wide-ranging 
applicability, eventually making it to the art of spreadsheet design.
Since then, Hermans and others catalog the odors of spreadsheets, 
often built off Fowler's (?) original specifications to articulate
the subtle problems in spreadsheet design. \par 
Collected, some of the intra-spreadsheet smells fall well within the
purview of this visualization tool. Some such smells can be found
in a figure [FIGURE]. \par 
The smells described above, then, entail a corresponding structure
visible in the visualization: the deeper a branch runs, or the
taller the range of arguments, the more it emits the stench of suboptimal
design. \par 
To be sure, these rank occurrences must be leveraged with their prevalence
across the dataset. One person's anomaly, though perhaps still interesting,
will not be as impactful as a smell of thousands. \par
Some examples are meaningful not for the space they take up but for how
they clash with the expectations around a function. For example, Excel
offers a number of functions that accept any number arguments given to 
them, up to a resource-defined limit. SUM and CONCATENATE are two such functions:
adding more arguments simply adds or appends that element in the same way
as every previous argument. Strange cases were found, then, in the cases where
these functions were used with only one argument. \par
Hermans, in her paper on the nuances of VLOOKUP and its common misuses, explore
the problem of the final, optional argument. A tool there helped, but how might
a visualization such as this illuminated the same?

@@

\subsection{Limitations}
Because the data collection depends on POI's formula parser, it affords no leeway or partial information from a formula; it either processes it perfectly or throws it out. As such, the visualization inhibits insight into anything with syntactical errors or third-party functions that can't be evaluated without additional tools. Note, however, that this does not include standard Excel errors like \#REF! and \#DIV/0, which the parser handles well. \par
Many functions have an expected order of arguments, but some, like SUM and MATCH, can be reordered in various ways and maintain their value. However, the representation that this tool uses reinforces for all functions that the order is important, even in these exceptions. \par
A larger and related problem is the loss of argument combination. In the example in section B, we saw how the tools conveys which arguments are used most often on each side of an equals sign, but it does not encode how these whether references were actually compared to strings and numbers the most or whether it corresponded more to all the arguments beneath those two

\section{Conclusion}

@@

-extensible to other spreadsheets?
% conference papers do not normally have an appendix

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%
%\end{thebibliography}

>>>>

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***
%                       ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf \usepackage[pdftex]{graphicx} \usepackage{wrapfig} % declare the
% path(s) where your graphic files are
\graphicspath{{pic/}} % and their extensions so you won't have to specify these
% with
% every instance of \includegraphics
\DeclareGraphicsExtensions{.png} \else % or other class option (dvipsone,
% dvipdf, if not using dvips). graphicx
% will default to the driver specified in the system graphics.cfg if no
% driver is specified.
% \usepackage[dvips]{graphicx}
% declare the path(s) where your graphic files are
% \graphicspath{{../eps/}}
% and their extensions so you won't have to specify these with
% every instance of \includegraphics
% \DeclareGraphicsExtensions{.eps}
\fi

@@

	\section{Case Study} The obvious question to pose to any visualization is this:
	Why does it matter that we see this way? In other words, it is not enough to
	prove that something can be visualized; instead we must also demonstrate what
	we can be gained from any particular visualization. As such, we raised a few
	tasks in the introduction in which we thought the tool could help: identifying
	places for API improvement, detecting bad smells or common programmer
	misunderstanding, and guiding spreadsheet education. To demonstrate the tool's
	applicability, then, we sought out places in the tree which best exemplify
	these respective concerns.

	\subsection{Bad Smells} Fowler's (?) seminal description of code smells
	underscored an important point of code quality: between perfect code and
	bug-crippled spaghetti,  there is a spectrum of code designs which, by
	themselves, are not faulty but nevertheless suggest problems and
	vulnerabilities in code design. Code smells, as we call them, soon commanded
	attention and earned wide-ranging applicability, eventually making it to the
	art of spreadsheet design. Since then, Hermans and others catalog the odors of
	spreadsheets, often built off Fowler's (?) original specifications to
	articulate the subtle problems in spreadsheet design. \par Collected, some of
	the intra-spreadsheet smells fall well within the purview of this visualization
	tool. Some such smells can be found in a figure [FIGURE]. \par The smells
	described above, then, entail a corresponding structure visible in the
	visualization: the deeper a branch runs, or the taller the range of arguments,
	the more it emits the stench of suboptimal design. \par To be sure, these rank
	occurrences must be leveraged with their prevalence across the dataset. One
	person's anomaly, though perhaps still interesting, will not be as impactful as
	a smell of thousands. \par Some examples are meaningful not for the space they
	take up but for how they clash with the expectations around a function. For
	example, Excel offers a number of functions that accept any number arguments
	given to them, up to a resource-defined limit. SUM and CONCATENATE are two such
	functions: adding more arguments simply adds or appends that element in the
	same way as every previous argument. Strange cases were found, then, in the
	cases where these functions were used with only one argument. \par Hermans, in
	her paper on the nuances of VLOOKUP and its common misuses, explore the problem
	of the final, optional argument. A tool there helped, but how might a
	visualization such as this illuminated the same?

@@

	\section{User Study} Midway through production, we conducted a brief,
	four-participant user study in order to gauge how well we were capturing the
	initial design goals. The study was, in general, a flexible and exploratory
	affair: rather than giving the users any specific tasks or pointed questions,
	we asked instead that they choose for themselves which trees to explore and
	report anything that they deemed interesting. Specifically, given trees rooted
	in the Enron dataset and at least 20 to thirty minutes, the participants
	situated themselves within the persona of a consultant asked to evaluate a
	company's spreadsheet practice, which could be arranged around questions like
	which functions on which the employees most depended, which sheets presented
	the most anomalous or dangerous designs, and so on. Interpretation of
	"spreadsheet practice" can certainly vary from participant to participant, but
	since we wanted a range of views and backgrounds despite the persona, we
	readily accepted this interpretive wiggle room. \par We clarified, furthermore,
	that these notes could be directed either at the quality of the data conveyed
	by the tool or at the tool's conveyance itself. Since the tool's philosophy was
	based in exploration, we decided that it would not have been appropriate to
	restrict what people were to look for, letting them find interesting stories of
	formula construction for themselves. Additionally, either kind of comment -- on
	data or on tool -- would direct us back to the central evaluation; a user's
	silence, or lack of any interesting findings, would tacitly indicate that
	perhaps the tool does not convey information well enough. \par Still, several
	limitations cropped up during the study that complicate the evaluation. For one
	thing, though the source of the data was not explicitly stated at the outset of
	the session, several users studied the file names (unchanged as they were) to
	deduce its roots in finance, if not Enron itself. Because of these notorious
	connotations, then, some were inclined from this moment of discovery to orient
	their exploration around finance-related functions, potentially limiting their
	findings or perceived users for the tool -- though some also said that, without
	this essential context, they could not have properly explored a dataset in the
	first place. Comments like these point out the trade-offs of unguided
	exploration, too: though it might not restrict them to a single purpose, it
	might, at the same time, not afford them any mindset at all with which to
	interpret the data. Another potential problem is that the data which users
	explored was processed and created before we had fully completed the tool,
	meaning some of the counts and patterns in the trees were inaccurate. However,
	we decided that this was not a pressing concern, being that we were primarily
	with whether any interesting reports could be made at all, not with whether
	these reports were 100\% accurate assessments of reality. \par Nevertheless,
	these user studies were a formative moment in evaluating the tool's
	performance, generating a number of directions or corrections in the process.
	We present an overview of responses below, with the four participants recorded
	simply as P1 through P4. \par \subsection{Positive Responses} The inclusion of
	formula examples, grounded in and linking back to the originating spreadsheets,
	proved to be an especial draw to the tool. After all, none of the participants
	professed themselves to be experts with Excel, their self-reported
	proficiencies ranging from "fairly familiar" (P2) to "not extremely familiar"
	(P4); even if they were experts, it would likely be unreasonable to expect an
	evenly distributed familiarity with the 100+ function trees available.
	Nevertheless, the list of actual formulas accessible through a node's
	double-click allowed the users a way to find examples of formulas with exactly
	the configuration they wanted. For example, as P1 explored the VLOOKUP
	function, even though they knew the official API more verbosely explained its
	signature, they said they could use the tool to collect more examples than the
	documentation offers. \par Furthermore, it offers a simple way of finding
	examples where certain functions were used together. though a standard
	text-search tool might yield the same results for a function alone,
	combinations can require more complex text queries, whereas here, the
	information is already captured in nodes. [This paragraph contrasts it with
	previous approaches, but lacks compelling evidence -- a lot of "mights"] \par
	\begin{wrapfigure}{l}{.12\textwidth} \centering \includegraphics{SUM}
		\label{"fig:sum"} \end{wrapfigure} The discovery of errors and bad design,
	whether by purpose or accident, produced the more jarring events of the study,
	and as such, most users commented on the tool's ability to locate these
	problems in spreadsheet design. This was often supported by the distinctive
	visual signatures of certain bad practices: multiple arguments produce towers
	of nodes, deep function nesting stretches the tree horizontally, and so on. As
	such, the participants remarked on the problems quickly after opening: P3,
	exploring the example for a 27-argument SUM (with SUM's initial appearance
	pictured left), diagnosed the situation with either incompetency or just plain
	weirdness; P2, meanwhile, decided that the spreadsheet's author "was having a
	bad day" when writing an AVERAGE with 18 distinct arguments. Aside from these
	implicit cues, there are also nodes which refer to formula errors, prefixed
	with \#s, which P4 said could be useful if marked well. Furthermore, once these
	errors are found, the participants could then open the offending spreadsheets,
	discovering that, even in context, some of these formulas remained
	incomprehensible, posing problems for mere functioning and maintainability.
	[Hermanns discussed in one of her paper the transfer scenarios -- would go well
	here.] The tool, then, aggregates these issues in a single place, accessible to
	either the unguided user and the ones who know exactly where to find these
	issues. \par The detection of bad design, however, is certainly not limited to
	a visualization tool, and some users remarked that a tool which printed a list
	of errors could suffice just as well for that purpose. However, they also
	remarked that this visual interface better supports the case when searching for
	new and unknown brands of bad smell.

	Because the exploratory design makes no qualitative judgment and visualizes
	benign designs with the problematic, it therefore allows for flexible
	interpretation of what is an issue; that is, a user might not realize a certain
	design is suboptimal until they see it in the tree and, at the point, find
	every instance of it.

	\subsection{Negative Responses} Some interaction with the tool underlined how a
	lack of contextual information may limit the tool. For example, the benefits
	described in the previous sections imply a hidden complication: if the tool
	educates through example but doesn't explicitly distinguish between good design
	and bad design, how can we be sure that people won't inadvertently learn bad
	design from it. Right now, we can't be sure; the tool would have to be

	supplemented with a distinguishing eye, informed on bad design already, and
	supplying the tool with this would take it beyond a design-agnostic view. [More
	on this in the conclusion.] \par Furthermore, just because a user can
	exhaustively explore every variation of how a function was used doesn't mean
	they will learn exactly what it does. P1, in their exploration, examined
	functions that they hadn't used before, such as KURT and PV, and described
	precisely what types of arguments it accepted and how many. However, even after
	exploring the spreadsheets, they could not accurately describe what the
	functions produced without consulting the official documentation. Whether
	through reticence of the tool or obscurity of the functions themselves, the
	exploratory environment is not enough for independent education. However,
	unlike the previous problem, this can be addressed, perhaps, by guiding the
	user directly to such documentation from within the tool without violating any
	design concerns. \par A threat to the exploratory philosophy may also mount
	from the problem of too much data, particularly in the popular functions, like
	SUM and IF. We had added some precautions before the study: for any position in
	the tree, for example, only the ten most common functions were displayed
	without clicking on the expansion arrow \includegraphics[scale=.35]{arrow}.
	Nevertheless, P3 remarked that they specifically avoided the most popular
	functions, anticipating a flood of nodes from which they could salvage nothing
	interesting -- which might not have been a bad prediction, considering that the
	SUM is built from 910 nodes and IF, at least 18000 [double-check these with
	refined data]. P2 corroborated this by explaining how they gravitated toward
	specific examples over the plane of nodes but complicates it further by also
	deeming the tool at its best when it shows either every possibility of an
	side-by-side (for comparison) or nothing on that branch at all, a balance that
	is hard to strike with node-dense trees. \par Throughout the study, it became
	apparent that some of the concepts, particularly optional arguments, were
	easily lost in the representation. To illustrate this, consider the following
	two trees: \par \includegraphics[width = .46\textwidth]{comparison} The
	difference is, perhaps, too subtle: on the left, the tree contains instances of
	AND functions with any number of arguments, up to four; on the right, the tree
	contains only instances with exactly four arguments. In other words, an
	instance of "=AND(A1, B2, C3, D4)" would be found in both trees, while
	"=AND(A1, B2)" would be found in only the left. Though the users naturally did
	not explicitly report misinterpreting the icons, it became clear in their
	out-loud thoughts that they viewed each box as a fundamentally different set of
	possibilities in formula construction rather than just a specific parameter.
	Though we clarified the meanings when these misunderstandings became apparent,
	these events point to a failure in the tool to properly indicate every function
	with clarity.   \par

	\section{Limitations} Beyond the negative responses in the user study, a few
	more inherent limitations beset the tool. These largely resulted from
	trade-offs in which we found no acceptable way to win the best of both worlds,
	and necessary compromises were made. \par Because the data collection depends
	on POI's formula parser, it affords no leeway or partial information from a
	formula; it either processes it perfectly or throws it out. As such, the
	visualization inhibits insight into anything with syntactical errors or
	third-party functions that can't be evaluated without additional tools. Note,
	however, that this does not include standard Excel errors like \#REF! and
	\#DIV/0, which the parser handles well. This poses more egregious problems,
	however, when POI does not support the parsing of certain functions, such as
	EOMONTH (as obscure as it may be), preventing the healthy growth of those
	trees. \par Many functions have an expected order of arguments, but some, like
	SUM and MATCH, can be reordered in various ways and maintain their value.
	However, the representation that this tool uses reinforces for all functions
	that the order is important, even in these exceptions. Future work could design
	a new family of tree to accommodate these loose cases, however, by bucketing
	order-insignificant arguments together; this, of course, would still have to
	address the profusion of co-presented data. \par A larger and related problem
	is the loss of argument combination. In the example in section B, we saw how
	the tools conveys which arguments are used most often on each side of an equals
	sign, but it does not encode how these whether references were actually
	compared to strings and numbers the most or whether it corresponded more to all
	the arguments beneath those two

	\section{Conclusion} Why design agnostic?

@@

	-extensible to other spreadsheets? % conference papers do not normally have an
	% appendix

@@

	% can use a bibliography generated by BibTeX as a .bbl file
	% BibTeX documentation can be easily obtained at:
	% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
	% The IEEEtran BibTeX style support page is at:
	% http://www.michaelshell.org/tex/ieeetran/bibtex/
	\bibliographystyle{IEEEtran} % argument is your BibTeX string definitions and
	% bibliography database(s)
	\bibliography{paper} %
	% <OR> manually copy in the resultant .bbl file
	% set second argument of \begin to the number of references
	% (used to reserve space for the reference number labels box)
	%\begin{thebibliography}{1}
	%
	%\bibitem{IEEEhowto:kopka}
	%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
	%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
	%
	%
	%\end{thebibliography}
